### 2021-07-08 rad-icfp-2021/unf.tex
.
fixing Sam's comments sec4

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar of Target UNF is given in Figure~\ref{fig:unf_source_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.

@@

We are now able to present a simpler transformation for purely functional reverse-mode from Source UNF to Target UNF.
The differentiation transformation is given in Figure~\ref{fig:diff_macro}.
To a list of types, the transformation adds a higher-order type. 
It is to be understood as the continuation for the pure storage of the gradient.
On primitive constants of Source UNF, it returns a pair. 
The first part is the initial term (pre-composed by a projection because it is not using the continuation variable).
The second part returns the continuation term pre-composed by the transpose Jacobian of the primitive.

@@

\begin{lemma}[Well typedness $\UNFSymbol$]
    Let !$\Gamma$=x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$ $\vdash$ e:B! be a term in Source. Then
    \begin{center}
        !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ $\UNFSymbol$(e): A$_{1}$,$\ldots$,A$_{n}$,B!.
    \end{center}
\end{lemma}

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of ANF transformation. 
$\UNFSymbol^{-1}$ does not preserve a typing property as nicely as $\UNFSymbol$, 
because it treats primitives with and without a $J$ differently.

>>>>

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar of Target UNF is given in Figure~\ref{fig:unf_target_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.

@@

We are now able to present a simpler transformation for purely functional reverse-mode from Source UNF to Target UNF.
The differentiation transformation is given in Figure~\ref{fig:diff_macro}.
To a list of types, the transformation adds a higher-order type. 
It is to be understood as the continuation for the pure storage of the gradient.
On primitive constants of Source UNF, it returns a pair. 
The first part is the initial term (pre-composed by a projection because it is not using the continuation variable).
The second part returns the continuation term pre-composed by the transpose Jacobian of the primitive. 
There is a sharp contrast between the simplicity of $\Dsynrevsymbol_{\rho}$ compared to $\directD{\rho}{\Gamma}{Y}$ from Figure~\ref{fig:direct_diff_macro}.

\begin{lemma}[Well typedness $\Dsynrevsymbol_{\rho}$]
    Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. \\
    Then 
    % \begin{center}
        !A$_{1}$,$\ldots$,A$_{n}$,A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\Dsynrevsymbol_{\rho}$(e): B$_{1}$,$\ldots$,B$_{m}$,B$_{1}\times\ldots\times$ B$_{m}$->$\rho$!
    % \end{center}
\end{lemma}
\begin{proof}
    By induction on derivation of !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$!.
\end{proof}

@@

\begin{lemma}[Well typedness $\UNFSymbol$]
    Let !$\Gamma$=x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$ $\vdash$ e:B! be a term in Source. Then
    \begin{center}
        !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ $\UNFSymbol$(e): A$_{1}$,$\ldots$,A$_{n}$,B!.
    \end{center}
\end{lemma}
\begin{proof}
    By induction on derivation of $\Gamma\vdash$!e:B!.
\end{proof}

@@

Next, the transformation from Target UNF to Target is presented in Figure~\ref{fig:unf_to_target}. 
We call this transformation $\UNFSymbol^{-1}$, but it is not a strict inverse of $\UNFSymbol$.
Doing $\UNFSymbol$ followed by $\UNFSymbol^{-1}$ performs some version of the ANF transformation \cite{sabry1993reasoning}. 
$\UNFSymbol^{-1}$ does not have as a simple typing property as $\UNFSymbol$, 
because it treats primitives with and without a $J$ differently and should be performed after $\Dsynrevsymbol_{\rho}$.

\begin{lemma}[Well typedness $\UNFSymbol^{-1}$]
    Let !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$! be a term in Source UNF. Then
    \begin{center}
        !x$_1$:A$_{1}$,$\ldots$,x$_n$:A$_{n}$,x$_{n+1}$:A$_{1}\times\ldots\times$ A$_{n}$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(e)): B$_m$ $\times$ (B$_{1}\times\ldots\times$ B$_{m}$->$\rho$) !.
    \end{center}
\end{lemma}
\begin{proof}
    By induction on derivation of !A$_{1}$,$\ldots$,A$_{n}$ $\vdash$ e: B$_{1}$,$\ldots$,B$_{m}$!.
\end{proof}
